---
title: "Feature-log-BF level curves for gene-set enrichment"
subtitle: "Targeting the SER per-feature log Bayes factor (localization evidence) over set size, enrichment, and per-gene signal"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    fig-format: png
engine: knitr
---

```{r}
#| label: setup
#| include: false
# knitr engine so the doc is polyglot (R + Python). Analysis chunks are Python via
# reticulate against the repo-root uv venv (one dir up from twogroup_experiments/).
# The analytic part needs only numpy/scipy/matplotlib; the confirmatory simulation
# loads the GO:BP design + gibss to fit the oracle SER.
proj_root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
knitr::opts_knit$set(root.dir = proj_root)
Sys.setenv(RETICULATE_PYTHON = file.path(proj_root, "..", ".venv", "bin", "python"))
library(reticulate)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This is the companion to `detectability_level_curves.qmd`. That notebook targets the
**expected enrichment LRT**; here we target the quantity that actually drives *localization*
in the enrichment SuSiE: the SER's per-feature log Bayes factor for the causal set,
`feature_log_bf` (what gibss calls `single_effects[l].feature_log_bf[causal]`).

Why switch: the LRT is a *profiled* likelihood-ratio statistic, but the SER reports a
*Bayes factor* - it integrates the enrichment coefficient over a prior instead of profiling
it. The two differ by an $O(\text{a few nats})$ penalty, so an LRT $= 25$ operating point only
buys `feature_log_bf` $\approx 10$, and the localization evidence
$\texttt{ser\_log\_bf}\approx\texttt{feature\_log\_bf}-\log p$ (with $p$ = number of gene sets)
is then near zero. Targeting `feature_log_bf` directly fixes this.

### The model, and the quadrature

`feature_log_bf` for the causal set is the **univariate** two-group log Bayes factor for that one
column: regress the z-scores on in-set membership and integrate the enrichment coefficient $\beta$
over its $\mathcal N(0,1)$ prior (the SER's prior). Out-of-set genes cancel between the numerator
and the null, so only the $m$ in-set genes enter, each with linear predictor $\eta = b_0 + \beta$.
An active gene's z is $g_1 = \mathcal N(\mu,\tau_0^2+1)$ (loc) or $\mathcal N(0,\sigma^2+1)$ (scale),
an inactive gene's is $g_0=\mathcal N(0,1)$, and $q_\beta(z) = (1-\sigma(b_0+\beta))g_0 + \sigma(b_0+\beta)g_1$.
For one data set,

$$ \texttt{feature\_log\_bf} \;=\; \log\!\int \exp\!\Big(\textstyle\sum_i \big[\log q_\beta(z_i)-\log q_0(z_i)\big]\Big)\,\mathcal N(\beta;0,1)\,d\beta . $$

**We compute $\mathbb E[\texttt{feature\_log\_bf}]$ directly, by quadrature - not with the
Wakefield/Laplace ABF.** The Wakefield approximation is built on the Wald statistic $z=\hat\beta/\mathrm{se}$,
which in logistic-type models is **non-monotone in the true effect** (the *Hauck-Donner* effect:
as the effect grows the standard error inflates faster than $\hat\beta$, so $z\to0$). That makes the
Wakefield $\mathbb E[\texttt{feature\_log\_bf}]$ spuriously *turn over* at large $b$ - larger
enrichment appearing to *lose* evidence. The Bayes-factor integral above has no such pathology, and
we evaluate its expectation exactly.

Write $S(\beta)=\sum_i d_\beta(z_i)$, $d_\beta(z)=\log q_\beta(z)-\log q_0(z)$. Because the $m$ genes
are iid, the vector $\big(S(\beta_k)\big)_k$ over a $\beta$-grid is, by the CLT,
$\mathcal N(M,C)$ with $M_k = m\,\mu_k$, $\mu_k=\mathbb E[d_{\beta_k}(Z)]$, and
$C_{kl}=m\,\mathrm{Cov}[d_{\beta_k}(Z),d_{\beta_l}(Z)]$ - all plain 1-D $z$-quadratures against the
data marginal $q_{\rm true}$. To second order in the fluctuations,

$$ \mathbb E[\texttt{feature\_log\_bf}] \;\approx\; \operatorname*{logsumexp}_k\!\big(M_k+\log\pi_k+\log\Delta\beta\big) \;+\; \tfrac12\Big(\textstyle\sum_k w_k C_{kk} - \sum_{kl} w_k w_l C_{kl}\Big), $$

with $\pi_k=\mathcal N(\beta_k;0,1)$ and posterior weights $w=\operatorname{softmax}(M+\log\pi+\log\Delta\beta)$.
The leading term is the annealed (plug-in) evidence; the correction accounts for finite-$m$
fluctuation of the log-integral. Both are deterministic quadratures - smooth, fast, and cached
across $m$.

Verified against Monte-Carlo `feature_log_bf`: the quadrature agrees to $\lesssim 1$ nat across the
whole $b$ range, **including the large-effect regime where the Wakefield ABF turns over**, and it is
monotone in $b$ and in $m$. The confirmatory section below reproduces this.

```{python}
#| label: core
#| code-summary: "Quadrature E[feature_log_bf] (exact Bayes-factor integral, no Wakefield)"
import numpy as np
from scipy.stats import norm
from scipy.special import logsumexp

B0 = -2.0                # intercept: log-odds of activity for an in-set gene under the null
TAU0 = 0.01              # loc per-gene effect sd (active gene z ~ N(mu, tau0^2 + 1))
C_PRIOR_SD = 1.0         # N(0, 1) prior on the enrichment coefficient beta (the SER's prior)
N_GENES = 19646          # rows of the gobp_10_500 design (only used to report set fractions)

# z grid (the observed marginal) and beta grid (the prior integral): two fixed quadratures.
_Z = np.linspace(-15.0, 25.0, 4500)
_DZ = _Z[1] - _Z[0]
_G0 = np.exp(-0.5 * _Z ** 2) / np.sqrt(2 * np.pi)          # inactive-gene density g0 = N(0,1)
K_BETA = 90
_BG = np.linspace(-3.0, 15.0, K_BETA)                      # beta nodes for the prior integral
_DBG = _BG[1] - _BG[0]
_LPI = norm.logpdf(_BG, 0.0, C_PRIOR_SD)


def _sigmoid(t):
    return 1.0 / (1.0 + np.exp(-t))


def _g1(scenario, s):
    """(mean, sd) of the active-gene z marginal g1 = N(mu, sd^2)."""
    return (float(s), np.sqrt(TAU0 ** 2 + 1.0)) if scenario == "loc" else (0.0, np.sqrt(s ** 2 + 1.0))


def _g1_pdf(scenario, s):
    mu, sd = _g1(scenario, s)
    return np.exp(-0.5 * ((_Z - mu) / sd) ** 2) / (sd * np.sqrt(2 * np.pi))


# feature_log_bf for the causal set is the *univariate* two-group log Bayes factor for that one
# column: regress z on in-set membership and integrate the enrichment coefficient beta over its
# N(0,1) prior. Out-of-set genes cancel, so only the m in-set genes enter, with linear predictor
# eta = b0 + beta. For one data set,
#   flb = log \int exp( sum_i [ log q_beta(z_i) - log q_0(z_i) ] ) N(beta; 0, 1) d beta,
#   q_beta(z) = (1 - sigma(b0 + beta)) g0(z) + sigma(b0 + beta) g1(z),   q_0 = q at beta = 0.
# We want E[flb] over the data (m iid genes ~ q_true, the beta = b marginal). The exponent
# S(beta) = sum_i d_beta(z_i) is a sum of m iid terms, so the vector (S(beta_k))_k is
# ~ Normal(m*mu, m*Sigma) by the CLT, mu_k = E[d_{beta_k}(Z)], Sigma_kl = Cov[d_{beta_k}, d_{beta_l}]
# - both plain z-quadratures. Then, to second order (Gaussian fluctuation of the log-integral),
#   E[flb] ~ logsumexp_k( M_k + log pi_k + log dbeta ) + 1/2 ( <C_kk>_w - <C_kl>_ww ),
# with M = m*mu, C = m*Sigma, and weights w = softmax(M + log pi + log dbeta).
# This is direct quadrature of the Bayes-factor integral - NO Wald / Wakefield z-score. It is
# therefore monotone in the effect and free of the Hauck-Donner pathology (the Wald z, hence the
# Wakefield ABF, turns over as the effect grows because the SE inflates faster than the estimate).
# Matches Monte-Carlo to <~1 nat across b, including the saturated large-effect regime.
_FLB_CACHE = {}


def _flb_ingredients(scenario, sig_axis, b0):
    """m-independent quadratures per (scenario, sig_axis, b0), cached across set sizes m. The data
    activity rate p = sigma(b0 + b) enters linearly (q_true = g0 + p (g1 - g0)), so mu and the
    second-moment matrices split into a g0 part and a (g1 - g0) part."""
    key = (scenario, b0, sig_axis.tobytes())
    if key in _FLB_CACHE:
        return _FLB_CACHE[key]
    wq = _sigmoid(b0 + _BG)                                          # (K,) membership prob per beta
    ny = len(sig_axis)
    aV = np.empty((ny, K_BETA)); bV = np.empty((ny, K_BETA))
    A = np.empty((ny, K_BETA, K_BETA)); Bm = np.empty((ny, K_BETA, K_BETA))
    dA = np.empty((ny, K_BETA)); dB = np.empty((ny, K_BETA))
    j0 = int(np.argmin(np.abs(_BG)))                                 # beta = 0 node
    for iy, s in enumerate(sig_axis):
        g1 = _g1_pdf(scenario, s)
        qb = (1 - wq)[:, None] * _G0[None, :] + wq[:, None] * g1[None, :]        # (K, nz)
        d = np.log(np.maximum(qb, 1e-300)) - np.log(np.maximum(qb[j0], 1e-300))[None, :]
        g0w = _G0 * _DZ; gdw = (g1 - _G0) * _DZ
        aV[iy] = (d * g0w[None, :]).sum(1); bV[iy] = (d * gdw[None, :]).sum(1)   # mu = aV + p bV
        A[iy] = (d * g0w[None, :]) @ d.T; Bm[iy] = (d * gdw[None, :]) @ d.T      # E[d d'] = A + p Bm
        dA[iy] = np.diag(A[iy]); dB[iy] = np.diag(Bm[iy])
    val = (aV, bV, A, Bm, dA, dB)
    _FLB_CACHE[key] = val
    return val


def e_flb_field(scenario, b_axis, sig_axis, m, b0=B0):
    """E[feature_log_bf] over enrichment b (x) and per-gene signal (y) for set size m. Deterministic
    quadrature (see note above); monotone in b and in m."""
    aV, bV, A, Bm, dA, dB = _flb_ingredients(scenario, sig_axis, b0)
    pt = _sigmoid(b0 + b_axis)                                        # (nb,) data activity rate
    out = np.empty((len(sig_axis), len(b_axis)))
    for iy in range(len(sig_axis)):
        MU = aV[iy][None, :] + pt[:, None] * bV[iy][None, :]          # (nb, K) per-gene mean mu(beta)
        M = m * MU
        logw = M + _LPI[None, :] + np.log(_DBG)
        lead = logsumexp(logw, axis=1)                               # (nb,) leading (annealed) term
        w = np.exp(logw - lead[:, None])                             # (nb, K) posterior weight on beta
        dC = m * (dA[iy][None, :] + pt[:, None] * dB[iy][None, :] - MU ** 2)      # diag(C)
        wA = w @ A[iy]; wB = w @ Bm[iy]
        quad = m * ((wA * w).sum(1) + pt * (wB * w).sum(1) - (w * MU).sum(1) ** 2)  # w' C w
        out[iy] = lead + 0.5 * ((w * dC).sum(1) - quad)              # + fluctuation correction
    return out


def e_flb_point(scenario, m, b, s, b0=B0):
    return float(e_flb_field(scenario, np.array([b]), np.array([s]), m, b0)[0, 0])


# Targets and shared config.
FLB_DETECT = 10.0        # "threshold detectability" (feature_log_bf ~ 10)
FLB_MID = 15.0           # intermediate operating point (feature_log_bf ~ 15)
FLB_LOCALIZE = 20.0      # localization (feature_log_bf ~ 20)
# (target value, linestyle, linewidth, legend label), drawn most- to least-evidence.
FLB_TARGET_STYLES = [(FLB_LOCALIZE, "-", 2.0, "localize (=20)"),
                     (FLB_MID, "--", 1.3, "intermediate (=15)"),
                     (FLB_DETECT, ":", 1.0, "detect (=10)")]
FLB_VMAX = 40.0          # colour saturates at 2x the localization target: the field runs to a few
                         # hundred, which would crush the detect/localize band (10-20) into the
                         # palest few percent of the colormap. Capping keeps that band legible.
SIZES = [50, 100, 200, 400]
SIZE_REFS = SIZES
REF_M = 200              # set size used for the colour fill in the b0-sweep panels
import matplotlib.pyplot as plt
print(f"N={N_GENES} genes, b0={B0}, tau0={TAU0}, prior beta ~ N(0,{C_PRIOR_SD:g}); "
      f"targets feature_log_bf = {FLB_DETECT:g} (detect) / {FLB_LOCALIZE:g} (localize)")
```

## The feature-log-BF landscape

We plot $\mathbb E[\texttt{feature\_log\_bf}]$ over enrichment $b$ and per-gene signal, for
**location** ($f_1=\mathcal N(\mu,\tau_0^2)$) and **scale** ($f_1=\mathcal N(0,\sigma^2)$).
`feature_log_bf` is *not* linear in the set size $m$, so there is no single per-gene field; instead
we give **one panel per set size** ($m = 50/100/200/400$, columns) for each scenario (rows). In each
panel the colour *is* that panel's own $\mathbb E[\texttt{feature\_log\_bf}]$ field, and the two
overlaid contours are that same field's **localization** ($=20$, solid) and **detection** ($=10$,
dotted) level curves - so the lines always trace the fill. The colour saturates at $40$ (the field
runs to a few hundred) to keep the $10$-$20$ band legible.

```{python}
#| label: compute
b_axis_loc = np.linspace(0.0, 4.0, 100)
mu_axis = np.linspace(0.0, 4.0, 100)
b_axis_scale = np.linspace(0.0, 4.0, 100)
sigma_axis = np.linspace(0.0, 6.0, 100)
_f200 = e_flb_field("loc", b_axis_loc, mu_axis, REF_M)
print("grid:", _f200.shape, "| E[flb] loc m=200 range:",
      (round(_f200.min(), 1), round(_f200.max(), 1)))
```

```{python}
#| label: fig-landscape
#| fig-width: 15
#| fig-height: 7.4
#| fig-cap: "E[feature_log_bf] by scenario (rows) and set size m (columns). In every panel the colour is that panel's own E[feature_log_bf] field (saturating at 40) and the overlaid crimson contours are its localization (=20, solid), intermediate (=15, dashed) and detection (=10, dotted) level curves. Deterministic quadrature of the Bayes-factor integral; b0=-2. Larger m reaches the targets at weaker b and weaker per-gene signal (curves move toward the origin left-to-right)."
_SCEN = [("loc", b_axis_loc, mu_axis, "per-gene shift  $\\mu$", "Location"),
         ("scale", b_axis_scale, sigma_axis, "per-gene effect sd  $\\sigma$", "Scale")]
LINE_C = "#b2182b"
fig, axes = plt.subplots(2, 4, figsize=(15, 7.4), sharex=True)
for r, (scen, b_axis, sig_axis, ylabel, rtitle) in enumerate(_SCEN):
    B, Y = np.meshgrid(b_axis, sig_axis)
    for c, m in enumerate(SIZES):
        ax = axes[r, c]
        fld = e_flb_field(scen, b_axis, sig_axis, m)
        cf = ax.contourf(B, Y, fld, levels=np.linspace(0, FLB_VMAX, 21), cmap="Blues", extend="max")
        for tval, ls, lw, _ in FLB_TARGET_STYLES:
            ax.contour(B, Y, fld, levels=[tval], colors=LINE_C, linewidths=lw, linestyles=ls)
        if r == 0:
            ax.set_title(f"m = {m}")
        if r == 1:
            ax.set_xlabel("enrichment log-odds  $b$")
        ax.set_ylabel(f"{rtitle}\n{ylabel}" if c == 0 else "")
        ax.set_box_aspect(1)
cb = fig.colorbar(cf, ax=axes, label="E[feature_log_bf]  (saturates at 40)", shrink=0.6, pad=0.02)
axes[0, 3].legend(handles=[plt.Line2D([], [], color=LINE_C, lw=lw, ls=ls, label=lab)
                           for _, ls, lw, lab in FLB_TARGET_STYLES],
                  loc="upper right", fontsize=7, frameon=True, framealpha=0.9)
fig
```

Reading left-to-right, the localization ($=20$) and detection ($=10$) curves both move toward the
origin as $m$ grows - a larger set reaches the same `feature_log_bf` at weaker $b$ and weaker
per-gene signal. The gap between the solid ($=20$) and dotted ($=10$) curves in each panel is the
*extra* signal localization costs over mere detection. Both curves are now monotone in $b$ (no
turn-back): the Hauck-Donner artifact of the Wakefield version is gone.

## Effect of the background rate $b_0$

```{python}
#| label: fig-b0-loc
#| fig-width: 18
#| fig-height: 4.4
#| fig-cap: "Location E[feature_log_bf] at m=200 over (b, mu) at four intercepts b0. Contours: that field's localization (=20, solid), intermediate (=15, dashed) and detection (=10, dotted) curves. Detectability peaks near b0~-1: too empty a background gives little contrast, too full and the null is already signal-rich."
B0_GRID = [-4, -3, -2, -1]


def _b0_row(scenario, b_axis, sig_axis, ylabel):
    # one set size (REF_M): colour is that field, contours are the same field's targets.
    B, Y = np.meshgrid(b_axis, sig_axis)
    fig, axes = plt.subplots(1, 4, figsize=(16, 4.6), sharex=True, sharey=True)
    for ax, b0 in zip(axes, B0_GRID):
        fld = e_flb_field(scenario, b_axis, sig_axis, REF_M, b0=b0)
        cf = ax.contourf(B, Y, fld, levels=np.linspace(0, FLB_VMAX, 21), cmap="Blues", extend="max")
        for tval, ls, lw, _ in FLB_TARGET_STYLES:
            ax.contour(B, Y, fld, levels=[tval], colors=LINE_C, linewidths=lw, linestyles=ls)
        ax.set(title=f"$b_0={b0}$ ($p_0={_sigmoid(b0):.2f}$)", xlabel="enrichment log-odds  $b$")
        ax.set_box_aspect(1)
    axes[0].set_ylabel(ylabel)
    fig.colorbar(cf, ax=axes, label=f"E[feature_log_bf] at $m={REF_M}$", shrink=0.85)
    return fig


_b0_row("loc", b_axis_loc, mu_axis, "per-gene shift  $\\mu$")
```

```{python}
#| label: fig-b0-scale
#| fig-width: 18
#| fig-height: 4.4
#| fig-cap: "Scale E[feature_log_bf] at m=200 over (b, sigma) at four intercepts b0. Contours: that field's localization (=20, solid), intermediate (=15, dashed) and detection (=10, dotted) curves."
_b0_row("scale", b_axis_scale, sigma_axis, "per-gene effect sd  $\\sigma$")
```

## Proposed simulation design points

For each set size $m$ and each target $T$ the design curve is the locus
$\mathbb E[\texttt{feature\_log\_bf}] = T$. We place **three points per curve** by normalized arc
length (enrichment-heavy / balanced / signal-heavy), on the detection ($T=10$), intermediate ($T=15$) and
localization ($T=20$) targets. That is $3\text{ regimes}\times 3\text{ targets} = 9$ points in each
of the $2\text{ scenarios}\times 4\text{ sizes} = 8$ panels, **72 points** total.

```{python}
#| label: design-compute
DESIGN_SIZES = [50, 100, 200, 400]
REGIME_ORDER = ["enrichment-heavy", "balanced", "signal-heavy"]
TARGETS = [("detect", FLB_DETECT), ("mid", FLB_MID), ("localize", FLB_LOCALIZE)]
_AXES = {"loc": (b_axis_loc, mu_axis), "scale": (b_axis_scale, sigma_axis)}


def tradeoff_points(scenario, b_axis, sig_axis, m, target):
    field = e_flb_field(scenario, b_axis, sig_axis, m)
    B, Y = np.meshgrid(b_axis, sig_axis)
    _f = plt.figure()
    segs = plt.contour(B, Y, field, levels=[target]).allsegs[0]
    plt.close(_f)
    if not segs:
        return []
    seg = max(segs, key=len)
    bw, sw = b_axis[-1] - b_axis[0], sig_axis[-1] - sig_axis[0]
    xn, yn = seg[:, 0] / bw, seg[:, 1] / sw
    d = np.concatenate([[0.0], np.cumsum(np.hypot(np.diff(xn), np.diff(yn)))])
    d /= d[-1]
    pts = [(float(np.interp(fr, d, seg[:, 0])), float(np.interp(fr, d, seg[:, 1]))) for fr in (0.0, 0.5, 1.0)]
    pts.sort(key=lambda p: p[1])
    return [{"m": m, "regime": r, "b": b, "signal": s} for r, (b, s) in zip(REGIME_ORDER, pts)]


design_all = []
for scen, (ba, sa) in _AXES.items():
    for tlabel, tval in TARGETS:
        for m in DESIGN_SIZES:
            for p in tradeoff_points(scen, ba, sa, m, tval):
                design_all.append(dict(p, scenario=scen, target=tlabel, target_val=tval))
print(f"{len(design_all)} design points (expect 72 = 2 scenarios x 4 sizes x 3 targets x 3 regimes)")
```

```{python}
#| label: fig-design
#| fig-width: 15
#| fig-height: 7.4
#| fig-cap: "Design points on the detection (=10, dotted line), intermediate (=15, dashed line) and localization (=20, solid line) curves, one panel per scenario x set size (72 points). Colour is that panel's E[feature_log_bf] field. Marker shape = regime (enrichment-heavy circle, balanced diamond, signal-heavy triangle); each marker sits on its target's contour."
REGIME_MARKER = {"enrichment-heavy": "o", "balanced": "D", "signal-heavy": "^"}
REGIME_COLOR = {"enrichment-heavy": "#E69F00", "balanced": "#CC79A7", "signal-heavy": "#009E73"}
_design_by = {}
for p in design_all:
    _design_by.setdefault((p["scenario"], p["m"]), []).append(p)
fig, axes = plt.subplots(2, 4, figsize=(15, 7.4), sharex=True)
for r, (scen, b_axis, sig_axis, ylabel, rtitle) in enumerate(_SCEN):
    B, Y = np.meshgrid(b_axis, sig_axis)
    for c, m in enumerate(SIZES):
        ax = axes[r, c]
        fld = e_flb_field(scen, b_axis, sig_axis, m)
        cf = ax.contourf(B, Y, fld, levels=np.linspace(0, FLB_VMAX, 21), cmap="Blues", extend="max")
        for tval, ls, lw, _ in FLB_TARGET_STYLES:
            ax.contour(B, Y, fld, levels=[tval], colors=LINE_C, linewidths=lw, linestyles=ls)
        for p in _design_by.get((scen, m), []):
            ax.plot(p["b"], p["signal"], marker=REGIME_MARKER[p["regime"]], ms=7.5,
                    mfc=REGIME_COLOR[p["regime"]], mec="white", mew=0.7, ls="none", zorder=5)
        if r == 0:
            ax.set_title(f"m = {m}")
        if r == 1:
            ax.set_xlabel("enrichment log-odds  $b$")
        ax.set_ylabel(f"{rtitle}\n{ylabel}" if c == 0 else "")
        ax.set_box_aspect(1)
fig.colorbar(cf, ax=axes, label="E[feature_log_bf]  (saturates at 40)", shrink=0.6, pad=0.02)
_leg = [plt.Line2D([], [], marker=REGIME_MARKER[r], color=REGIME_COLOR[r], mec="white", ls="none", ms=8, label=r)
        for r in REGIME_ORDER]
_leg += [plt.Line2D([], [], color=LINE_C, lw=lw, ls=ls, label=lab) for _, ls, lw, lab in FLB_TARGET_STYLES]
axes[0, 3].legend(handles=_leg, loc="upper right", fontsize=6.5, frameon=True, framealpha=0.9)
fig
```

## Confirmatory simulation

`feature_log_bf` for the causal set is the *univariate* two-group log Bayes factor for that one
column, so we don't fit the full SuSiE over all $p$ sets: we simulate the in-set z-scores and
compute the log-BF directly (integrate the enrichment coefficient over its $\mathcal N(0,1)$
prior). We Monte-Carlo **all 72 design points** and compare to the quadrature $\mathbb E[\texttt{feature\_log\_bf}]$.

```{python}
#| label: sim-table
#| code-summary: "Monte-Carlo of the univariate feature_log_bf at all 48 design points vs the quadrature"
from scipy.stats import norm
from scipy.special import logsumexp


def _mc_flb(scenario, m, b, s, n_rep=300, b0=B0):
    """Monte-Carlo univariate two-group log-BF for the causal set (in-set genes only)."""
    g1m, g1sd = _g1(scenario, s)
    bg = np.linspace(-2.0, 12.0, 200); lpi = norm.logpdf(bg, 0, 1); ldb = np.log(bg[1] - bg[0])
    w = _sigmoid(b0 + bg); w0 = _sigmoid(b0)
    out = np.empty(n_rep)
    for r in range(n_rep):
        rng = np.random.default_rng(r)
        active = rng.binomial(1, _sigmoid(b0 + b), m)
        z = np.where(active == 1, rng.normal(g1m, g1sd, m), rng.normal(0, 1, m))
        d0, d1 = norm.pdf(z, 0, 1), norm.pdf(z, g1m, g1sd)
        ll = np.sum(np.log((1 - w)[:, None] * d0[None, :] + w[:, None] * d1[None, :]), axis=1)
        ll0 = np.sum(np.log((1 - w0) * d0 + w0 * d1))
        out[r] = logsumexp(ll - ll0 + lpi + ldb)          # integrate b over N(0,1)
    return out


_TLABEL = {"detect": ("detection", FLB_DETECT), "mid": ("intermediate", FLB_MID), "localize": ("localization", FLB_LOCALIZE)}
_hdr = ["scenario", "m", "regime", "b", "mu", "sigma", "target", "analytic E[flb]", "MC flb mean", "MC q25", "MC q75"]
_lines = []
for _tkey in ("detect", "mid", "localize"):
    _name, _tval = _TLABEL[_tkey]
    _rows = sorted([p for p in design_all if p["target"] == _tkey],
                   key=lambda p: (p["scenario"], p["m"], REGIME_ORDER.index(p["regime"])))
    _lines += [f"Target feature_log_bf ~ {_tval:g} ({_name}); N={N_GENES}, b0={B0:g}, 300 reps, {len(_rows)} points:\n",
               "| " + " | ".join(_hdr) + " |", "| " + " | ".join("---" for _ in _hdr) + " |"]
    for _p in _rows:
        _sc, _m, _b, _s = _p["scenario"], _p["m"], _p["b"], _p["signal"]
        _samp = _mc_flb(_sc, _m, _b, _s); _q = np.quantile(_samp, [0.25, 0.75])
        _mu, _sg = (_s, TAU0) if _sc == "loc" else (0.0, _s)
        _lines.append("| " + " | ".join([_sc, str(_m), _p["regime"], f"{_b:.2f}", f"{_mu:.2f}", f"{_sg:.2f}",
            f"{_tval:g}", f"{e_flb_point(_sc, _m, _b, _s):.1f}", f"{_samp.mean():.1f}",
            f"{_q[0]:.1f}", f"{_q[1]:.1f}"]) + " |")
    _lines.append("")
SIM_MD = "\n".join(_lines)
```

```{r}
#| label: sim-table-emit
#| echo: false
#| output: asis
# The Python `asis` stdout stream drops silently this late in the render (reticulate
# capture interacts with the many matplotlib figures above), so we emit the table string
# from R instead, which is reliable.
cat(reticulate::py$SIM_MD)
```

**Columns.** *regime* - which arc-length point on the target curve (enrichment-heavy / balanced /
signal-heavy); *b / mu / sigma* - enrichment log-odds and $f_1=\mathcal N(\mu,\sigma^2)$ at that
point; *target* - the `feature_log_bf` the point sits on; *analytic E[flb]* - the quadrature (hits
the target by design); *MC flb mean / q25 / q75* - the simulated univariate `feature_log_bf` over
reps. Every one of the 72 points is Monte-Carlo verified. The localization evidence follows as
$\texttt{ser\_log\_bf}\approx\texttt{feature\_log\_bf}-\log p$ ($\log p\approx 8.6$): $\approx 1.4$
at target 10, $\approx 6.4$ at target 15, $\approx 11.4$ at target 20.

## Takeaways

- **Compute the Bayes factor by quadrature, not Wakefield.** `feature_log_bf` is the SER's per-set
  evidence. Its expectation is a direct quadrature of the Bayes-factor integral (CLT over genes +
  a second-order fluctuation term), so we can build iso-`feature_log_bf` level curves and pick a
  target with a real margin above the $\log p\approx 8.6$ multiplicity floor.
- **The Wakefield/Wald ABF is wrong here (Hauck-Donner).** Because it is built on the Wald z-score,
  the Wakefield expectation is non-monotone in the effect - it turns over at large $b$, so larger
  enrichment appears to *lose* evidence. The quadrature is monotone in $b$ and $m$ and matches
  Monte-Carlo to $\lesssim 1$ nat across the whole range, saturation included.
- **Detection $\ne$ localization.** At `feature_log_bf` $=10$ the set is barely detected
  (ser_log_bf $\approx 1$, low PIP); at $=20$ it localizes (ser_log_bf $\approx 11$, high PIP).
  The solid ($=20$) and dotted ($=10$) curves in each panel make the extra cost explicit.
- **One panel per set size.** Because `feature_log_bf` is not linear in $m$, each panel shows its
  own field with its own contours - no cross-$m$ overlay on a mismatched fill.
```
