---
title: "GO:BP operating points: method performance across the detectability landscape"
subtitle: "Enrichment-SER methods over set size, localization-evidence strength, and enrichment-vs-signal regime"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    fig-format: png
engine: knitr
---

```{r}
#| label: setup
#| include: false
# knitr engine so the doc is polyglot; analysis chunks are Python via reticulate against
# the repo-root uv venv (one dir up from twogroup_experiments/). Same pattern as the
# other analysis/*.qmd notebooks.
proj_root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
knitr::opts_knit$set(root.dir = proj_root)
Sys.setenv(RETICULATE_PYTHON = file.path(proj_root, "..", ".venv", "bin", "python"))
library(reticulate)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dev = "png", dpi = 110)
```

## What this is

The 011 GO:BP operating-points experiment (`analysis/feature_logbf_level_curves.qmd`)
simulates enrichment on the GO:BP design (GOBP_ sets of `c5.all`, size in [10,500], fixed
background log-odds `b0=-2`), placing each cell on an **iso-`E[feature_log_bf]` curve** - a
matched *localization-evidence* target for the oracle two-group SER. The design cube, per
signal-type (`loc`: shifted-mean $f_1=\mathcal N(\mu,\tau_0^2)$; `scale`: heavier
$f_1=\mathcal N(0,\sigma^2)$):

- **set size** $m \in \{50,100,200,400\}$;
- **strength** (evidence tier) $\in$ {weak ($\approx 10$), intermediate ($\approx 15$),
  strong ($\approx 20$)} - `ser_log_bf` $\approx$ `feature_log_bf` $- \log p$ ($\log p\approx 8.6$),
  so weak barely detects, strong localizes;
- **regime** (arc-length position on the iso-evidence curve) $\in$ {enrich (high enrichment
  log-odds, weak per-gene signal), balanced, signal (strong per-gene signal, weak enrichment)}.

**Layout.** There are too many axes to overlay at once, so each figure is a grid with
**rows = strength, cols = regime**, one figure per **(set size $m$, signal-type)**; $m$ and
signal-type are paged across figures. Methods are overlaid within each panel. The grid is
sized to the tiers actually on disk, so this **pilot** run (only `loc`, $m=100$, weak+strong)
renders as **2×3**, and the full sweep fills in to **3×3** with no change to the notebook -
switch by editing the one `SC` line below.

```{python}
#| label: load
#| code-summary: "Load the supercollection's cells, methods, and pip/cs reductions"
import sys
sys.path.insert(0, "analysis")
import matplotlib.pyplot as plt
import polars as pl
import op_landscape as ol

# >>> switch pilot <-> full here <<<
SC = "011-gobp-operating-points"     # or "011-pilot"

CELLS = ol.cells(SC)
MM = ol.method_meta(SC)
PIP = ol.load_reduction("pip", SC)
CS = ol.load_reduction("cs", SC)
CS_PN = ol.load_reduction("cs", SC, include_null=True)   # + paired nulls, for the detection ROC
AVAIL = ol.available(SC).rows()   # [(scenario, m), ...] present on disk

print(f"supercollection: {SC}")
print(f"cells on disk:   {len(CELLS)}  ->  (scenario, m) figures: {AVAIL}")
print(f"methods:         {ol.methods(SC)}")

# Methods legend table (built here as a string; emitted from R below to dodge the
# reticulate `asis` stdout dropout in figure-heavy renders).
_mt = MM.select("method_display", "is_oracle", "is_thresholded", "threshold").sort(
    ["is_oracle", "is_thresholded", "method_display"], descending=[True, False, False])
_lines = ["| method | role | censoring threshold |", "| --- | --- | --- |"]
for r in _mt.iter_rows(named=True):
    role = "oracle" if r["is_oracle"] else ("thresholded" if r["is_thresholded"] else "model")
    thr = "-" if r["threshold"] is None else f"{r['threshold']:.0f}"
    _lines.append(f"| {r['method_display']} | {role} | {thr} |")
METHODS_TABLE = "\n".join(_lines)
```

**Methods (display name, role).**

```{r}
#| label: methods-table
#| echo: false
#| output: asis
# The table string is built in Python (chunk `load`); emit it from R, which is reliable
# where the Python `asis` stdout stream drops silently in figure-heavy reticulate renders.
cat(reticulate::py$METHODS_TABLE)
```

Colour encodes method family (twogroup Okabe-Ito palette); the cox censoring threshold is
encoded as line style (`@1` solid, `@2` dashed, `@3` dash-dot, `@4` dotted).

## Power vs FDP (PIP selection)

Per cell, the parametric power-vs-FDP curve as the PIP inclusion threshold sweeps high→low.
Top-left = an efficient ordering of PIPs. Rows = strength (weak→strong), cols = regime.

```{python}
#| label: fig-power-fdp
#| fig-width: 11
#| fig-height: 8
for scen, m in AVAIL:
    ol.figure_power_fdp(SC, scen, m, PIP, MM)
    plt.show()
```

## Coverage vs credible-set size (localization)

Empirical coverage of the causal set vs **median** CS size, swept over the CS β-grid (CS size
is heavy-tailed over ~5k overlapping sets, so median not mean). Marked points are the CS at the
**0.95 (circle), 0.80 (square), 0.50 (diamond)** nominal levels; each marker's horizontal
position vs the dashed line at its nominal reads as calibration (on the line = calibrated, left
= anti-conservative). Lower curve at a given coverage = tighter (more efficient) localization.

```{python}
#| label: fig-coverage-size
#| fig-width: 11
#| fig-height: 8
for scen, m in AVAIL:
    ol.figure_coverage_size(SC, scen, m, CS, MM)
    plt.show()
```

**Credible-set size by nominal level.** One row per cell and method; a coverage and a
size column for each of the 0.95 / 0.80 / 0.50 nominal levels (the marked points above).

```{python}
#| label: covsize-table-build
#| include: false
COVSIZE_TABLE = "\n\n".join(ol.coverage_size_table_md(SC, scen, m, CS, MM) for scen, m in AVAIL)
```

```{r}
#| label: covsize-table
#| echo: false
#| output: asis
cat(reticulate::py$COVSIZE_TABLE)
```

**Columns.** *strength / regime* - the cell (rows/cols of the figure); *method* - the SER
method; *cov 95/80/50* - fraction of the 50 replicates whose causal set falls inside the
β-CS at that nominal (calibrated ⇒ ≈ β); *size 95/80/50* - the β-CS size (number of gene-sets)
as `median [q25, q75]` over the replicates.

## Localization ROC (coverage vs CS footprint)

The coverage-vs-size curve recast as an ROC with an absolute reference. Sweeping the CS level β:
**y = coverage** of the causal set (`P(causal ∈ β-CS)`), **x = the CS footprint** = mean CS size
as a fraction of all ~5k gene-sets. A random set of footprint `f` covers the causal set with
probability `f`, so the dotted **y = x diagonal is exact chance** - a method whose credible set is
no more informative than a random set that big rides the diagonal. Genuine localization bows to
the **top-left** (high coverage at tiny footprint). The marked point is the nominal 95% CS. Rows =
strength, cols = regime, all methods overlaid; one figure per (signal-type, set size).

```{python}
#| label: fig-localization-roc
#| fig-width: 11
#| fig-height: 9
for scen, m in AVAIL:
    ol.figure_localization_roc(SC, scen, m, CS, MM)
    plt.show()
```

## PIP calibration

Every calibration plot uses the same encoding and method ordering: rows = regime, one
method per panel in its method colour, dots sized by #gene-sets in the PIP bin (the PIP
saturates - almost every set at PIP≈0, a few at ≈1 - so the big extreme dots are where
calibration lives), dotted diagonal = perfect calibration (below = overconfident). Two method
subsets throughout: the **cross-method** comparison (twogroup oracle/loc/scale, linear,
cox-reversed, cox@2, logistic@2) and the **threshold sweep** (cox/logistic at tau = 1..4).

### Spotlight — cross-method (pooled over strength + set size)

```{python}
#| label: fig-cal-spotlight-cross
#| fig-width: 13
#| fig-height: 6
for scen in sorted({s for s, _ in AVAIL}):
    ol.figure_pip_calibration_spotlight(SC, scen, PIP, MM, methods=ol.CROSS_METHODS)
    plt.show()
```

### Spotlight — threshold sweep (cox/logistic @ tau = 1..4, pooled)

```{python}
#| label: fig-cal-spotlight-thresh
#| fig-width: 14
#| fig-height: 6
for scen in sorted({s for s, _ in AVAIL}):
    ol.figure_pip_calibration_spotlight(SC, scen, PIP, MM, methods=ol.THRESHOLD_METHODS)
    plt.show()
```

### Full facet — cross-method (per set size, aggregated over strength)

```{python}
#| label: fig-cal-facet-cross
#| fig-width: 13
#| fig-height: 6
for scen, m in AVAIL:
    ol.figure_pip_calibration(SC, scen, m, PIP, MM, methods=ol.CROSS_METHODS)
    plt.show()
```

### Full facet — threshold sweep (per set size, aggregated over strength)

```{python}
#| label: fig-cal-facet-thresh
#| fig-width: 14
#| fig-height: 6
for scen, m in AVAIL:
    ol.figure_pip_calibration(SC, scen, m, PIP, MM, methods=ol.THRESHOLD_METHODS)
    plt.show()
```

## Log-BF detection ROC (enriched vs null)

Is the SER log Bayes factor a trustworthy **detection** statistic? Per cell, positive = an
enriched replicate, negative = its paired `null_b0` replicate (same design + signal, enrichment
$b=0$); score = `ser_log_bf`; methods overlaid. Broken out across the landscape (rows = strength,
cols = regime) - the point being to see detection sharpen as strength grows and vary by regime.
The diagonal is chance.

```{python}
#| label: fig-detection-roc
#| fig-width: 11
#| fig-height: 8
for scen, m in AVAIL:
    ol.figure_detection_roc(SC, scen, m, CS_PN, MM)
    plt.show()
```

## Landscape summary

Birds-eye view the per-cell grids bury: one small **m × strength** heatmap per method (pooled
over regime) of the median 95%-CS size - the "method performance across the detectability
landscape" the title names. Darker = tighter localization. (Pilot: one $m$ row, weak/strong;
fills to the full 4×3 landscape with the full run.)

```{python}
#| label: fig-landscape
#| fig-width: 12
#| fig-height: 5
for scen in sorted({s for s, _ in AVAIL}):
    ol.figure_landscape_heatmap(SC, CS, MM, scenario=scen, stat="cs_size")
    plt.show()
```
