---
title: "GO:BP operating points: credible-set summary tables"
subtitle: "Compact per-method CS metrics over regime × set size (pooled over strength)"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
engine: knitr
---

```{r}
#| label: setup
#| include: false
proj_root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
knitr::opts_knit$set(root.dir = proj_root)
Sys.setenv(RETICULATE_PYTHON = file.path(proj_root, "..", ".venv", "bin", "python"))
library(reticulate)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## What this is

Compact credible-set summaries for the 011 GO:BP operating-points experiment - the
scannable companion to the long per-cell tables in the landscape notebook. Every table:

- **rows = method** (the 7 cross-methods: twogroup oracle/loc/scale, linear, cox-reversed,
  cox@2, logistic@2);
- **columns = a 2-level header, regime (outer) × set size m (inner)**;
- **cells = one metric at the 95% CS, at the page's strength tier** (each cell aggregates that
  tier's replicates for the regime × size);
- pages run over `(signal-type × strength)`; within a page, one table per metric.

The design is 1 causal set among p = 5,268 GO:BP sets, so a credible set is a needle-in-haystack
localizer. A CS is "**small**" if it has ≤ 10 gene-sets. Five metrics per scenario:

- **CS coverage** - `P(causal ∈ 95%-CS)`; the CS's contract (well-behaved ⇒ ≈ 0.95).
- **median CS size** - the cost: how many gene-sets the CS carries.
- **small CS power** - `P(causal ∈ CS AND |CS| ≤ 10)`: the truth is caught *and* pinned to a
  short list (a genuinely useful call, not a diffuse set that merely contains it).
- **small CS coverage** - `P(causal ∈ CS | |CS| ≤ 10)`: *given* a small CS, how often is it
  right - can you trust a small CS when you get one.
- **small CS median size** - median `|CS|` among the small (≤10) CSs: how tight the small ones are.

Then, below a divider, the **null** block (paired `null_b0`, b=0 - no causal set). Coverage is
undefined there, but a small CS on a null *is* a false confident localization, so:

- **null small-CS rate** - `P(|CS| ≤ 10 | null)`; the false confident-localization rate (want ≈ 0).
- **null median CS size** - a well-behaved null CS should stay diffuse (large ⇒ appropriately uncertain).
- **null small-CS median size** - median `|CS|` on the false-positive subset (|CS| ≤ 10).

In the causal tables the **best method per column is bold** and any cell that **under-covers**
(coverage < 0.93) is **shaded light red** - so tight/high-power cells that aren't calibrated pop
out. In the null tables red instead marks a **null small-CS rate > 0.05** (too many false calls).

```{python}
#| label: load
#| code-summary: "Load reductions"
import sys
sys.path.insert(0, "analysis")
import op_landscape as ol

SC = "011-gobp-operating-points"        # or "011-pilot"
MM = ol.method_meta(SC)
CS = ol.load_reduction("cs", SC, include_null=True)   # causal + paired-null cells
SCENARIOS = sorted({s for s, _ in ol.available(SC).rows()})

CAUSAL_KEYS = ["coverage", "median_size", "loc_power", "localized_coverage", "small_median_size"]
NULL_KEYS = ["null_small_rate", "null_median_size"]
STRENGTH_ORDER = ["weak", "intermediate", "strong"]
_HR = "<hr style='margin:22px 0;border:0;border-top:2px solid #999'>"

def _block(scen, strength):
    causal = "\n\n".join(
        ol.cs_summary_html(SC, scen, CS, MM, metric=k, strength=strength, is_null=False)
        for k in CAUSAL_KEYS)
    nulls = "\n\n".join(
        ol.cs_summary_html(SC, scen, CS, MM, metric=k, strength=strength, is_null=True,
                           shade_high=(0.05 if k == "null_small_rate" else None))
        for k in NULL_KEYS)
    return causal + "\n\n" + _HR + "\n\n" + nulls

def _present_strengths(scen):
    return [s for s in STRENGTH_ORDER
            if any(c["strength"] == s for c in ol.cells(SC) if c["scenario"] == scen)]

# One markdown+HTML string, paged by scenario then strength; each "## scen - strength" is a
# real heading (into the TOC), followed by that page's 5 causal + 2 null tables. Blank lines
# around every heading and raw-HTML block so Pandoc keeps them separate.
_parts = []
for scen in SCENARIOS:
    for st in _present_strengths(scen):
        _parts.append(f"## {scen} - {st}\n\n{_block(scen, st)}\n")
ALL_TABLES = "\n\n".join(_parts)
print("built pages:", [(s, st) for s in SCENARIOS for st in _present_strengths(s)])
```

One page per **(signal-type × strength)**. Each page has the five causal tables (**CS coverage**,
**median CS size**, **small CS power**, **small CS coverage**, **small CS median size**), a
divider, then the two null tables (**null small-CS rate**, **null median CS size**).

```{r}
#| label: tbl-all
#| echo: false
#| output: asis
cat(reticulate::py$ALL_TABLES, "\n\n", sep = "")
```
